D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-13 ootavacuous.tex
Tighten wording for non-trivial semantic dependencies
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
shows an example that builds on the multiplication-by-zero
discussion in
Section~\ref{sec:Semantic Dependencies Are Per Execution}.
Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
\co{x} and \co{y} on lines~9 and~10, respectively.
Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
the value that line~15 loads from \co{z}.
Because all three variables are initialized to zero, changes to the value
loaded from \co{x} cannot affect the value stored by line~11 to \co{z}
unless there is also a change loaded from \co{y} and vice versa.
However, if both \co{x} and \co{y} change, the value stored to \co{z}
will change.

Therefore, as noted in
Section~\ref{sec:Semantic Dependencies Are Per Execution},
there is no semantic dependency from the load from \co{x} to the
store to \co{z}, nor is there a semantic dependency from the load
from \co{y} to the store to \co{z}.
However, there \emph{is} a semantic dependency from the \emph{combination}
of the loads from \co{x} and \co{y} to the store to \co{z}.

@@

For further evidence that there is no semantic dependency from \co{y}
to \co{z} when \co{x} is zero, see the equivalent\footnote{
	But equivalent only because the atomics are all non-volatile!}
program shown in
Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
Here, there is not even a load from \co{y} when \co{x} is zero, so there
cannot possibly be a semantic dependency involving this non-existent load.
A similar transformation would load \co{x} only if \co{y} was non-zero,
emphasizing the lack of any semantic dependency between \co{x} and \co{y}.

In both
Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
an OOTA cycle would need to affect both the \co{x} and the \co{y} branches
of this compound semantic dependency.
However, we again conjecture that tools designed to find OOTA cycles could
satisfy themselves with having found either of the two resulting cycles.

@@

Because the values loaded from \co{x} and \co{y} on lines~9 and~10
are non-zero, changes to either \co{x} or \co{y} will affect the
value stored to \co{z} by line~11.
There is therefore a separate semantic dependency from the load from
\co{x} to the store to \co{z} and from the load from \co{y} to the store
to \co{z}, which matches typical definitions of semantic dependency.
Further
Yet
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
are merely two different executions of the same program.

>>>>

Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
shows an example that builds on the multiplication-by-zero
discussion in
Section~\ref{sec:Semantic Dependencies Are Per Execution}.
Line~11 of \co{P0()} stores to \co{z} the product of the values loaded from
\co{x} and \co{y} on lines~9 and~10, respectively.
Similarly, lines~16 and~17 store to \co{x} and \co{y}, respectively,
the value that line~15 loads from \co{z}.
Because all three variables are initialized to zero, and because loads
can be ordered before any store, there is an execution in which all
three loads return the value zero, in which case all three stores will
of course store the value zero.
In any related execution where the value loaded from \co{y} remains zero,
non-zero values loaded from \co{x} cannot affect the value stored by
line~11 to \co{z}.
Only executions in which differ in the values loaded from both \co{x} and
\co{y} can cause the value stored to \co{z} to change to a non-zero value.

Therefore, as noted in
Section~\ref{sec:Semantic Dependencies Are Per Execution},
in an execution where the loads from both \co{x} and \co{y} return zero,
there is no semantic dependency from the load from \co{x} to the
store to \co{z}, nor is there a semantic dependency from the load
from \co{y} to the store to \co{z}.
However, there \emph{is} a semantic dependency from the \emph{combination}
of the loads from \co{x} and \co{y} to the store to \co{z}.

@@

For further evidence that there is no semantic dependency from \co{y}
to \co{z} in executions where the load from \co{x} is zero, see the
equivalent\footnote{
	But equivalent only because the atomics are all non-volatile!}
program shown in
Listing~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero}.
Here, there is not even a load from \co{y} in executions where the
load from \co{x} returns zero, so there cannot possibly be a semantic
dependency involving this non-existent load.
A similar transformation would load \co{x} only in executions where the
load from \co{y} returned non-zero, emphasizing the lack of any semantic
dependency between \co{x} and \co{z}.

In both
Listings~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and~\ref{lst:OOTA Conditional Multiplication Example, Initial Value Zero},
an OOTA cycle (such as the one indicated by their \co{exists} clauses)
would need to affect both the \co{x} and the \co{y} components of this
compound semantic dependency.
However, we again conjecture that tools designed to find OOTA cycles could
satisfy themselves with having found either of the two resulting cycles.

@@

Starting with the intuitive execution where the values loaded from \co{x}
and \co{y} on lines~9 and~10 are non-zero, executions that differ only
in the value loaded from either \co{x} or \co{y} will still affect the
value stored to \co{z} by line~11.
There is therefore a separate semantic dependency from the load from
\co{x} to the store to \co{z} and from the load from \co{y} to the store
to \co{z}, which matches typical definitions of semantic dependency.
Yet
Listing~\ref{lst:OOTA Multiplication Example, Initial Value Zero}
and
Listing~\ref{lst:OOTA Multiplication Example, Initial Value 1}
are merely two different executions of the same program.
